\section{Background and Related Work}
\label{sec:background}

\paragraph{OV circuit.} For a standard decoder attention head, the value and
output projections compose into a single rank-$d_\text{head}$ map
$W_{OV} = W_V W_O \in \mathbb{R}^{d_\text{model}\times d_\text{model}}$,
which acts on the source token's residual stream and adds the result to the
destination token's residual stream \citep{elhage2021mathematical}. The SVD
$W_{OV} = U\Sigma V^\top$ factorizes this map into read-in directions $V$ and
write-out directions $U$ scaled by singular values $\sigma_i$.

\paragraph{Helical and circular features.}
\citet{engels2024not} show that models encode numerical concepts (days of the
week, months, integers) in approximate circles or helices within the residual
stream; probing such structure typically involves training a linear probe on
cached activations. Our work differs in that compasses are read directly from
the weight matrix, without a probe and without supervised training.

\paragraph{Linear feature discovery.} DAS \citep{geiger2024das}, the linear
representation hypothesis \citep{park2024linearrep}, and the geometry of truth
\citep{marks2024geometry} all seek causal low-dimensional features in
activation space. These methods search over the \emph{residual stream}
directly; our compasses localize such features to a \emph{specific
attention head}, giving a sparser description with the same causal semantics.

\paragraph{Function vectors and rank-1 interventions.} Function vectors
\citep{todd2024function} and ROME-style rank-1 edits \citep{meng2022rome}
identify single directions that transport task information. Our compasses
generalize this picture to a $2$D plane on which every angle is a potential
steering direction, giving a continuous family of interventions rather than a
pair of antipodal ones.
