\section{Discussion}
\label{sec:discussion}

\paragraph{What a compass is and is not.} A semantic compass is a causal
feature localized to an attention head: its plane is a $2$D subspace of
the head's write-out space, and injections along it produce a continuous
family of behavior changes parameterized by an angle. A compass is
\emph{not} a claim that the head exclusively computes this axis. Higher
singular directions in the same head will carry other features; the
compass picks out one coherent $2$D slice in which a categorical axis is
readable and steerable.

\paragraph{Why OV-SVD works as a discovery device.} In a transformer
block, $W_{OV}$ determines the set of directions the head can add to the
residual stream. Its top SVD directions are therefore the candidate
``output languages'' of the head. When a pair of such directions
corresponds to a categorical axis (e.g., masculine vs.\ feminine), it
must correspond to \emph{something the head actually writes}, not just
something we can correlate post-hoc. This is the structural reason
OV-SVD is a stronger discovery method than probing the residual stream
directly: the feature is tied to a specific weight matrix from the
start.

\paragraph{Why polar shape differs without implying different
mechanisms.} Section~\ref{sec:shape} summarized four axes on which the
polar plots can differ (amplitude, singular-value scale, phase, neutral
offset). None of them change the underlying fact that the plane is
causal and sinusoidally-responsive; they only rescale, rotate, and
translate the picture. Future work should standardize polar plots by
dividing radii by $A(\alpha)$ so that the three models overlay.

\paragraph{Four-way dials: resolved in \S\ref{sec:harmonic}.} The two-harmonic
fit of a four-token probe distinguishes three distinct plane topologies
that are indistinguishable from decode-table inspection alone. Phi-3
L24H17 is a genuine four-way dial ($A_2 / A_1 = 0.32$), while L24H28
is two overlapping antipodal axes ($A_2 / A_1 \leq 0.06$). A natural
direction for future work is to sweep all $(u_i, u_j)$ pairs in a
model and classify each plane's dial topology automatically.

\paragraph{Limitations.}
\begin{enumerate}
    \item \textbf{Prompt dependence.} Baseline logit separations depend
    on our choice of three neutral / plus / minus prompts. We chose
    short, ambiguous prompts to minimize baseline bias; robustness to
    prompt paraphrase remains to be quantified.
    \item \textbf{Storage vs.\ steering locality.} Our head-ablation
    study (\S\ref{sec:headshare}) shows that in Gemma-2-2B the compass
    plane is both storage-local ($14\%$ of signal, $13.5\times$ over
    any other head) \emph{and} steering-local, whereas in GPT-2 and
    Phi-3 the compass is steering-local but not storage-local: the
    signal is distributed redundantly across heads. This is a
    substantive scope clarification: compass identifies where a rank-$2$
    causal push lands, not necessarily where the model keeps the
    category when no push is applied.
    \item \textbf{Tokenizer constraints.} Phi-3 tokenizes leading-space
    forms as multi-token; we therefore used bare tokens (\texttt{month},
    \texttt{hour}) which may subtly shift the logit-difference target.
    \item \textbf{Limited axis breadth.} We demonstrate gender in three
    models and temporal/entity in one. Sentiment, tense, formality,
    and animacy remain untested.
\end{enumerate}

\paragraph{Future work.} A natural next step is automated compass
discovery: scan all $(\ell, h, i, j)$ tuples and rank by the joint score
\emph{amplitude-linearity in $\alpha$ $\times$ phase-stability $\times$
vocabulary-decoding coherence}. Such a search could surface compasses
we currently find by reading decode tables manually. A second direction
is using compasses for lightweight steering: a rank-$2$ angular rotation
in one head is a $2 d_\text{model}$-parameter edit, far sparser than
full-vector patching.

\paragraph{Conclusion.} The SVD of an attention head's OV matrix is
not just a diagnostic object: its top singular planes routinely hold
causal, low-dimensional semantic features that can be read off directly
from the weights. The resulting \emph{semantic compass} is a simple,
architecture-agnostic primitive for mechanistic interpretability that
complements both circuit-level and feature-level methods.
