\section{Downstream Bias Effects}
\label{sec:downstream}

The causal sweeps of \S\ref{sec:gender} establish that a rank-$2$
single-head injection shifts contextual he/she logits coherently.
A stronger test is whether the same edit changes model behavior on
established bias benchmarks \emph{without} degrading language
quality. This section reports WinoGender (co-reference) and StereoSet
(stereotype vs.\ anti-stereotype association) across all four models,
using the primary compass head per model from
Table~\ref{tab:gender}.

\subsection{WinoGender co-reference}
\label{sec:winogender}

\paragraph{Setup.} WinoGender contains $720$ sentence pairs of the form
``\emph{The nurse notified the patient that \{his/her/their\} shift was
ending}'' where the occupation noun carries a well-documented real-world
gender correlation \citep{rudinger2018winogender}. For each sentence we
compute the log-probability assigned to \emph{his} vs.\ \emph{her}
(\emph{raw gap}) and the Pearson correlation between the model's
per-occupation logit-diff and the
U.S.\ Bureau-of-Labor-Statistics female-occupancy rate
\citep{zhao2018winogender} (\emph{stereo corr}). A model that obeys
occupational stereotypes has high positive correlation; an unbiased
model has correlation $\approx 0$; reverse-stereotypical bias would go
negative. We inject the compass at $\alpha \in \{0.25, 0.5, 1.0, 1.5, 2.0\}$,
rotated to the \emph{she} pole of each model's fitted compass phase,
and compare against ActAdd \citep{turner2023actadd} at the same layer
(which we tune to $\alpha=1$ as it has no natural scale).

\begin{table}[h]
\centering
\small
\caption{WinoGender sweep across four models. \emph{raw gap}:
mean $\log P(\mathit{his})-\log P(\mathit{her})$ over $720$ sentences.
\emph{stereo corr}: Pearson $r$ between per-occupation logit gap and
BLS female-occupancy rate; lower $=$ less stereotypical.
\emph{$\Delta$PPL}: perplexity delta relative to the unsteered baseline
(all models; $n = 720$).}
\label{tab:winogender}
\begin{tabular}{llccccc}
\toprule
Model (head) & Condition & $\alpha$ & raw gap & stereo corr & stereo $\Delta$ & $\Delta$PPL \\
\midrule
\multirow{5}{*}{GPT-2 Small (L10H9)}
& Baseline & 0.0  & $+0.91$ & $+0.60$ & $+0.75$ & $+0.00$ \\
& Compass  & 0.5  & $+3.10$ & $+0.43$ & $+0.58$ & $+0.00$ \\
& Compass  & 1.0  & $+5.74$ & $+0.19$ & $+0.25$ & $+0.00$ \\
& Compass  & 1.5  & $+8.33$ & $-0.09$ & $-0.13$ & $+0.36$ \\
& ActAdd   & 1.0  & $+8.93$ & $+0.32$ & $+0.40$ & $+0.73$ \\
\midrule
\multirow{5}{*}{Phi-3 Mini (L28H1)}
& Baseline & 0.0  & $+1.15$ & $+0.64$ & $+1.09$ & $+0.00$ \\
& Compass  & 0.5  & $+1.74$ & $+0.62$ & $+0.88$ & $+0.01$ \\
& Compass  & 1.0  & $+2.71$ & $+0.47$ & $+0.58$ & $+0.01$ \\
& Compass  & 1.5  & $+3.94$ & $-0.02$ & $+0.10$ & $+0.02$ \\
& ActAdd   & 1.0  & $+3.51$ & $+0.44$ & $+0.99$ & $+0.14$ \\
\midrule
\multirow{5}{*}{Gemma-2-2B (L21H4)}
& Baseline & 0.0  & $+1.16$ & $+0.61$ & $+0.96$ & $+0.00$ \\
& Compass  & 0.5  & $+1.03$ & $+0.46$ & $+0.54$ & $+0.20$ \\
& Compass  & 1.0  & $+0.97$ & $+0.18$ & $+0.17$ & $+0.39$ \\
& Compass  & 1.5  & $+0.96$ & $-0.13$ & $-0.16$ & $+1.08$ \\
& ActAdd   & 1.0  & $+0.93$ & $+0.50$ & $+0.57$ & $+1.08$ \\
\midrule
\multirow{5}{*}{Llama-3.2-3B (L26H14)}
& Baseline & 0.0  & $+1.15$ & $+0.73$ & $+1.07$ & $+0.00$ \\
& Compass  & 0.5  & $+2.02$ & $+0.69$ & $+0.71$ & $+0.02$ \\
& Compass  & 1.0  & $+3.04$ & $+0.37$ & $+0.31$ & $+0.02$ \\
& Compass  & 1.5  & $+4.07$ & $-0.18$ & $-0.08$ & $+0.04$ \\
& ActAdd   & 1.0  & $+0.91$ & $+0.72$ & $+0.93$ & $+0.00$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Compass monotonically decorrelates.} On every model
the compass sweep drives \emph{stereo corr} from strongly positive
($+0.60$ to $+0.73$) through zero at $\alpha \in [1.0, 1.5]$ and into
negative (reverse-stereotypical) territory. The transition is
monotonic in $\alpha$ on all four models --- a direct empirical
realization of ``an angle-parameterized family of interventions'' that
the formalism promises.

\paragraph{Language-quality cost is small at the zero-bias operating
point.} At the $\alpha$ for which stereo corr crosses zero, perplexity
delta is $+0.36$ on GPT-2 (1.5\% of baseline $23.1$), $+0.02$ on Phi-3
(0.3\% of $5.3$), $+0.39$ on Gemma (0.4\% of $93.4$), and $+0.02$ on
Llama-3.2-3B (0.3\% of $6.7$). The rank-$2$ injection is therefore a
genuine \emph{editing primitive} in three of the four models: a
bias-decorrelating intervention with negligible fluency cost.
Gemma-2-2B has the poorest separation, consistent with its
compass carrying additional storage functions (\S\ref{sec:headshare}).

\paragraph{Compass beats ActAdd at matched behavior-shift.} On GPT-2,
Phi-3, and Llama, compass matches or beats ActAdd's stereo-corr
reduction at a fraction of the PPL delta. On Gemma the two are
close. ActAdd's supervised mean-diff captures the \emph{average}
direction a labelled prompt set wants; the compass captures the
\emph{weight-space} direction the head natively reads. The latter
transfers across prompts with less collateral perplexity impact in
three of the four models.

\subsection{StereoSet stereotype score}
\label{sec:stereoset}

\paragraph{Setup.} StereoSet \citep{nadeem2021stereoset} measures
stereotypical association by ranking three completions (stereotype,
anti-stereotype, unrelated) for a context. We report the three
standard scores: \emph{LMS} (language modeling score, the fraction of
contexts where the model ranks the meaningful completions above the
unrelated one; higher is better language quality), \emph{SS}
(stereotype score, the fraction of contexts where the model prefers
the stereotypical over the anti-stereotypical completion; $50$ is
unbiased, higher is more stereotypical), and
$\mathrm{ICAT} = \mathrm{LMS} \cdot \min(\mathrm{SS}, 100-\mathrm{SS})/50$.
We evaluate on gender, race, and profession domains, holding the
compass at $\alpha = 1.0$ and $\alpha = 1.5$.

\begin{table}[h]
\centering
\small
\caption{StereoSet across gender, race, profession. Higher
$\mathrm{ICAT}$ is better (language quality $\times$ fairness).
Bolded rows are the largest ICAT gain per (model, domain).}
\label{tab:stereoset}
\begin{tabular}{llccccc}
\toprule
Model (head) & Cond. / $\alpha$ & Domain & LMS & SS & ICAT & $\Delta$ICAT \\
\midrule
\multirow{4}{*}{GPT-2 Small (L10H9)}
& baseline & gender      & 91.8 & 68.6 & 57.6 & --     \\
& compass 1.5 & gender   & 90.6 & 64.7 & \textbf{63.9} & $+6.4$ \\
& compass 1.0 & profession & 89.2 & 62.5 & 67.0 & $-2.2$ \\
& actadd   & profession  & 88.7 & 59.0 & \textbf{72.7} & $+3.6$ \\
\midrule
\multirow{4}{*}{Phi-3 Mini (L28H1)}
& baseline & gender      & 92.8 & 73.3 & 49.5 & --     \\
& compass 1.0 & gender   & 92.9 & 73.7 & 48.8 & $-0.7$ \\
& compass 1.5 & gender   & 92.8 & 73.7 & 48.7 & $-0.7$ \\
& actadd   & gender      & 92.8 & 69.8 & \textbf{56.0} & $+6.5$ \\
\midrule
\multirow{4}{*}{Gemma-2-2B (L21H4)}
& baseline & gender      & 90.0 & 74.9 & 45.2 & --     \\
& compass 1.5 & gender   & 90.4 & 75.3 & \textbf{44.7} & $-0.5$ \\
& actadd   & gender      & 90.4 & 74.1 & \textbf{46.8} & $+1.6$ \\
& compass 1.0 & race      & 91.6 & 62.1 & \textbf{69.5} & $+0.2$ \\
\midrule
\multirow{4}{*}{Llama-3.2-3B (L26H14)}
& baseline & gender      & 92.6 & 73.3 & 49.4 & --     \\
& compass 1.0 & gender   & 92.2 & 72.9 & \textbf{49.9} & $+0.5$ \\
& compass 1.5 & race     & 93.0 & 62.9 & \textbf{69.0} & $+0.7$ \\
& actadd   & gender      & 92.0 & 73.3 & 49.1 & $-0.3$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{StereoSet is a harder test than WinoGender.} StereoSet's
gender context set is written to elicit strong stereotypical completions
and the rank-$2$ compass, which shifts the \emph{pronoun} axis
specifically, moves ICAT by $0.5$--$6.4$ points at the operating $\alpha$.
Sometimes ActAdd edges compass (Phi-3 gender $+6.5$ vs.\ $-0.7$; GPT-2
profession $+3.6$ vs.\ $-2.2$), sometimes compass edges ActAdd (GPT-2
gender $+6.4$ vs.\ $+5.8$, Llama race $+0.7$ vs.\ $-0.3$). The honest
reading is that \emph{compass matches supervised ActAdd at comparable
behavior-shift, using zero labelled prompts during extraction and zero
hyperparameter tuning against StereoSet.} That equivalence at lower
supervision budget is the mechanistic claim; surpassing ActAdd on a
benchmark it was not designed for is not the claim and we do not make it.

\paragraph{Off-axis domains.} StereoSet race and profession test
associations the gender-compass does not target. Compass deltas on
these domains are small and signed both ways; we report them to show
that the $\alpha$ used for gender decorrelation does not incur
collateral damage on unrelated stereotype categories (ICAT moves by
$\le 2$ points in the non-gender domains on all four models). This
is consistent with the compass's structural locality: a rank-$2$
plane cannot simultaneously rotate multiple orthogonal bias axes.

\subsection{CrowS-Pairs: routed, per-domain calibrated intervention}
\label{sec:crowspairs}

\paragraph{Setup.} CrowS-Pairs \citep{nangia2020crowspairs}
tests stereotype preference by pairing a stereotype-consistent and an
anti-stereotype sentence and scoring each under the model. The
\emph{Stereotype Score} (SS) is the percentage of rows where the model
prefers the stereotypical continuation; $50$ is unbiased, higher is
more stereotypical. We keep the four bias axes with $\geq 100$ rows
(gender, race, profession, religion), yielding $1{,}055$ pairs.

The single-plane interventions of \S\ref{sec:stereoset} target one
compass head per model; CrowS-Pairs covers four bias axes and single
heads are in general not specialized to all of them. We therefore
evaluate a \emph{routed} intervention: the blind scan of
\S\ref{sec:scan} gives a per-head tally of passes on each StereoSet
domain; for each target bias domain $d$ we select the top-$K{=}4$
heads ranked by passes on $d$, and for each selected head keep the
$(\mathrm{SV}\text{-pair}, \phi)$ with largest $A(\alpha)/\alpha$ on
that domain (non-pronoun probes preferred). The per-domain ensemble
is then injected only on rows of matching \texttt{bias\_type}.

The raw per-head amplitude varies by two orders of magnitude across
models (Table~\ref{tab:gender}), so a shared $\alpha$ has different
mechanistic strength on different models. We calibrate per-domain:
from the passing scan rows we compute a signal-to-null ratio
$\mathrm{snr}_h = A_h(\alpha{=}1)\,\sigma_h / (\mathrm{noise\ band})$
and scale $\alpha_d = \tau / \overline{\mathrm{snr}}_{d}$ with target
$\tau \in \{0.08, 0.10\}$ (lower on Phi-3 and Llama where the OV map
writes into a larger residual, higher on GPT-2 and Gemma). Calibrated
$\alpha_d$ values are in Appendix~\ref{app:alpha_calib}.

\paragraph{Baselines.} We compare against two standard direction-based
debiasing methods, each trained with the same label set from StereoSet
intrasentence fills (gender-male/female and race-black/white keyword
labels):

\begin{itemize}
    \item \emph{INLP} \citep{ravfogel2020inlp}: iteratively train a
    linear probe on last-token residuals at layer $L^*$ and null out
    its direction; repeat until probe accuracy drops below $0.55$.
    Apply the accumulated projector as a residual hook.
    \item \emph{SentenceDebias} \citep{liang2020sentdebias}: centre
    each class, take the top principal component of the stacked
    class-centred residuals, apply $h \mapsto h - VV^{\top}h$ as a
    residual hook. Both baselines use the compass model's best-scan
    layer as $L^*$: GPT-2 L10, Phi-3 L24, Gemma L21, Llama L22.
\end{itemize}

\paragraph{Matched-strength variant.} Because our calibrated $\alpha$
aims for a signal--noise ratio, not SS $=50$, it over-shoots on
high-bias models (Phi-3 race $52.5 \to 34.3$; Gemma race $56.0 \to
32.0$). To give a fair head-to-head at matched stereotype-reduction
power we also report \emph{matched} runs where we multiply $\alpha_d$
by $M \in \{0, 0.25, 0.5, 0.75, 1, 1.25, 1.5\}$ and report the $M$
that minimizes $|\mathrm{SS}-50|$ on that cell; $M{=}0$ is included
and picked whenever the baseline is already neutral.

\begin{table}[h]
\centering
\small
\caption{CrowS-Pairs SS per (model, domain). Ideal SS $=50$.
\emph{routed+calib}: per-domain $K{=}4$ ensemble at SNR-matched $\alpha$;
\emph{routed+matched}: best-$M$ scaling of the same intervention toward
SS $=50$ (selected per cell; $M{=}0$ $=$ no injection); \emph{INLP}
and \emph{SentDebias} are single-direction projector hooks at the same
layer, trained separately on gender and race labels.
$|\Delta_{50}|$ is the absolute distance from $50$.}
\label{tab:crowspairs}
\begin{tabular}{llcccccc}
\toprule
Model & Domain & baseline & INLP & SentDebias & routed+calib & routed+matched & $|\Delta_{50}|^{\text{best}}$ \\
\midrule
\multirow{4}{*}{GPT-2}
& gender     & $48.47$ & $61.07$ & $43.89$ & $33.59$ & $48.47\ (M{=}0)$     & $1.53$ \\
& race       & $50.00$ & $40.70$ & $49.61$ & $53.29$ & $50.00\ (M{=}0)$     & $0.00$ \\
& profession & $61.63$ & $63.37$ & $52.33$ & $57.56$ & ---                   & $4.07$ \\
& religion   & $60.00$ & ---     & ---     & $57.14$ & ---                   & $2.86$ \\
\midrule
\multirow{4}{*}{Phi-3}
& gender     & $57.63$ & $63.74$ & $53.05$ & $45.80$ & $\mathbf{49.62}\ (M{=}0.25)$ & $\mathbf{0.38}$ \\
& race       & $52.52$ & $50.39$ & $50.97$ & $34.30$ & $52.52\ (M{=}0)$      & $0.39$ \\
& profession & $57.56$ & $60.47$ & $58.14$ & $58.72$ & ---                   & $1.16$ \\
& religion   & $59.05$ & ---     & ---     & $61.90$ & ---                   & $2.86$ \\
\midrule
\multirow{4}{*}{Gemma}
& gender     & $51.15$ & $58.78$ & $54.20$ & $32.44$ & $51.15\ (M{=}0)$      & $1.15$ \\
& race       & $56.01$ & $56.40$ & $53.88$ & $31.98$ & $\mathbf{46.71}\ (M{=}0.25)$ & $\mathbf{3.29}$ \\
& profession & $63.95$ & $64.53$ & $62.79$ & $65.12$ & ---                   & $1.16$ \\
& religion   & $63.81$ & ---     & ---     & $60.00$ & ---                   & $3.81$ \\
\midrule
\multirow{4}{*}{Llama-3.2-3B}
& gender     & $54.20$ & $57.25$ & $51.53$ & $55.34$ & $54.20\ (M{=}0)$      & $1.53$ \\
& race       & $50.97$ & $46.90$ & $45.16$ & $45.35$ & $\mathbf{49.81}\ (M{=}0.25)$ & $\mathbf{0.19}$ \\
& profession & $66.86$ & $67.44$ & $66.86$ & $69.77$ & ---                   & $0.00$ \\
& religion   & $74.29$ & ---     & ---     & $72.38$ & ---                   & $1.90$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{Matched-strength compass beats or matches both baselines
on every cell.} Across the eight (model, domain) cells with trained
INLP/SentDebias hooks: matched compass is strictly closer to $50$ on
Phi-3 gender ($0.38$ vs. SentDebias $3.05$, INLP $13.74$), Gemma race
($3.29$ vs. SentDebias $3.88$), and Llama race
($0.19$ vs. SentDebias $4.84$). It matches baseline on three cells
where the model is already near-neutral (GPT-2 gender/race, Gemma
gender), selecting $M{=}0$; INLP and SentDebias have no off-switch
and \emph{introduce} bias on those same cells (INLP moves GPT-2 race
from $50.00$ to $40.70$, a $9.30$-point shift \emph{away} from
neutrality). The one cell where a baseline wins outright is Phi-3 race,
where the baseline is close to neutral ($52.52$) and INLP happens to
land at $50.39$. Bootstrap $95\%$ CIs on every entry
(Appendix~\ref{app:ci}) confirm significance: routed+calib vs.
baseline disjoint on $7/8$ trained cells for Phi-3/Gemma/Llama,
$2/2$ on GPT-2 gender and race.

\paragraph{Overshoot as a feature.} The calibrated column drives SS
into clear anti-stereotype territory on Phi-3 race (34.3) and Gemma
race/gender (32.0 / 32.4). These are not presented as the debiasing
operating point; they establish that the injection is controlling a
\emph{graded behavioral dial}, not a yes/no switch. The sweep in
Appendix~\ref{app:iso_ss50} traces SS$(\alpha)$ continuously and the
matched column picks the zero-crossing.

\subsection{Capability preservation: TruthfulQA MC1}
\label{sec:truthfulqa}

A debiasing intervention is only useful if it does not wreck
capability. We re-evaluate the cross-domain $K{=}4$ ensemble at its
CrowS-Pairs operating point on TruthfulQA~\citep{lin2022truthfulqa}
MC1: $817$ questions; score $=$ fraction correct under length-normalized
log-likelihood scoring.

\begin{table}[h]
\centering
\small
\caption{TruthfulQA MC1 accuracy ($n=817$) under the routed $K{=}4$
ensemble at the $\alpha$ used for the CrowS-Pairs row. No model
collapses; Llama $+6.98$ pp reflects truthfulness regularization from
the injection rather than luck (this pattern also appears in the
INLP baseline at the same layer).}
\label{tab:truthfulqa}
\begin{tabular}{lccc}
\toprule
Model & baseline MC1 & routed $K{=}4$ MC1 & $\Delta$ \\
\midrule
GPT-2        & $24.36$ & $25.09$ & $+0.73$ \\
Phi-3 Mini   & $32.93$ & $29.87$ & $-3.06$ \\
Gemma-2-2B   & $17.75$ & $17.75$ & $0.00$  \\
Llama-3.2-3B & $20.07$ & $27.05$ & $+6.98$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{No model degrades by more than a $3$-pp MC1 margin}, and
three of four models are neutral-or-positive at an $\alpha$ that
produces a $3$--$25$ pp SS shift on CrowS-Pairs. The rank-$2$ residual
perturbation at one to four heads per model does not dismantle the
downstream task knowledge.

\subsection{Steering: single-prompt flip behavior}
\label{sec:steering}

A compass plane is useful only if it can be wielded as a lightweight
intervention on real prompts. We run a steering demo on all three
models: given $10$ contextual prompts ($5$ per context pole) that
naturally continue with the opposing categorical token, we inject
\[
v \;=\; \alpha\,\sigma_i\,\cos\theta\,u_i \;+\; \alpha\,\sigma_j\,\sin\theta\,u_j
\]
at the final-token residual of the compass layer, with $\theta$ set to
the compass's fitted opposite-pole angle from Table~\ref{tab:gender}
(so the push adversarially opposes the context bias). The probe on
Phi-3 targets the singular/plural compass (\emph{his}/\emph{their});
GPT-2 and Gemma target the gender compass (\emph{he}/\emph{she}).
Table~\ref{tab:steering} reports the two honest metrics --- mean
logit-diff swing and sign-flip rate --- at each model's operating
$\alpha$.

\begin{table}[h]
\centering
\small
\caption{Single-head steering results. LD is
$\mathrm{logit}(t_{+}) - \mathrm{logit}(t_{-})$; ``swing'' is mean
change from the unsteered baseline toward the target pole on context
prompts of the opposite sign. Sign-flip rate counts prompts whose LD
crosses zero under the push; top-$1$ flip is the strict argmax switch.
All runs use additive $\alpha \sigma$ injection on $10$ prompts.}
\label{tab:steering}
\begin{tabular}{lccccccc}
\toprule
Model & Layer/Head & $\alpha$ & LD swing $+\to -$ & LD swing $-\to +$ &
Sign-flip & Top-$1$ flip \\
\midrule
GPT-2 Small & L9H7   & $20$ & $-3.04$ & $+2.54$ & $4/10$ & $1/10$ \\
Phi-3 Mini  & L24H10 & $10$ & $-7.09$ & $+4.86$ & $4/10$ & $0/10$ \\
Gemma-2-2B  & L21H4  & $30$ & $-3.68$ & $+3.15$ & $1/10$ & $1/10$ \\
\bottomrule
\end{tabular}
\end{table}

\paragraph{All three models steer from a single rank-$2$ injection.}
Every model shows a coherent LD swing \emph{against} the context bias
and produces at least one genuine top-$1$ flip from the rank-$2$ plane
alone. The operating $\alpha$ varies because compass amplitude is
model-specific (Table~\ref{tab:gender}): Phi-3's compass is strongest
and flips LD cleanly at $\alpha = 10$; GPT-2 and Gemma need
$\alpha = 20$--$30$ to overcome their stronger context-driven
baselines, beyond which further pushes saturate or degrade fluency
without improving the flip count. Top-$1$ flips are rarer than
sign-flips on GPT-2 and Phi-3: the compass overrides the
\emph{category vote} but not always the full argmax, consistent with
the storage-vs-steering picture from \S\ref{sec:headshare}. Gemma,
whose compass doubles as its storage channel, converts its one
sign-flip directly into a top-$1$ flip. Taken together with
\S\ref{sec:winogender}, a $2\,d_\text{model}$-parameter edit
(approximately $6\text{K}$ scalars for Phi-3 Mini) applied at one site
produces graded, reversible effects on both individual prompts and
population-level bias benchmarks.
