\section{Introduction}
\label{sec:intro}

Mechanistic interpretability aims to explain what individual components of a
neural language model compute. Two dominant methodologies have emerged:
\emph{circuit analysis} \citep{elhage2021mathematical,wang2023interpretability,conmy2023acdc}, which
identifies subsets of attention heads and MLPs responsible for a specific
behavior, and \emph{linear probing / distributed alignment}
\citep{park2024linearrep,geiger2024das,marks2024geometry}, which searches for
low-dimensional causal features inside the residual stream. The first has
component-level granularity but seldom describes what a circuit
\emph{represents}; the second has feature-level granularity but seldom
localizes the feature to a specific weight matrix.

We show that these two goals meet naturally in the SVD of a single attention
head's output-value matrix. Let $W_{OV}^{(l,h)} = W_V^{(l,h)} W_O^{(l,h)}
\in \mathbb{R}^{d_\text{model} \times d_\text{model}}$ be the compressed
per-head OV map, and let $W_{OV}^{(l,h)} = U\,\Sigma\,V^\top$ be its SVD.
Pairs of column directions $(u_i, u_j)$ of $U$ span two-dimensional planes
in the output (residual-stream) space. We call such a plane a \emph{semantic
compass} when (i) vocabulary projection onto the plane yields a smooth
angular ordering of semantically related tokens and (ii) residual-stream
injection along $\cos\theta\,u_i + \sin\theta\,u_j$ shifts the model's next-token
logits sinusoidally as a function of $\theta$, with phase stable across
injection scales.

\paragraph{Contributions.}
\begin{enumerate}
    \item \textbf{Compass definition and test battery.} We formalize the
    semantic compass as a low-rank OV projection and describe a nine-test
    battery---decoding, cyclicity, permutation, random-plane, principal-angles,
    scan-planes, semantic-ablate, causal-patch, self-test---that isolates
    causal compasses from geometric accidents (\S\ref{sec:methods}).
    \item \textbf{Four-model replication for gender.} We identify
    gender-pronoun compasses in GPT-2 Small (L9H7, L10H9), Phi-3 Mini
    (L24H10, L28H1), Gemma-2-2B (L21H4), and Llama-3.2-3B (L26H14), and
    show that a causal $\alpha$-sweep produces a sinusoidal response whose
    amplitude scales linearly with $\alpha$ in every model
    (\S\ref{sec:gender}). A blind causal scan of all top-$4$ SV-pairs in
    each model ($864$--$4032$ planes per model) recovers the same heads
    without using decoded vocabulary, with per-head concentration
    significance of $p < 10^{-6}$ against a full-OV null
    (\S\ref{sec:scan}).
    \item \textbf{Two non-pronoun compasses in Phi-3.} We surface a temporal-unit
    compass (\emph{month} vs.\ \emph{hour}, L24H17) and an entity-type compass
    (\emph{item} vs.\ \emph{operation}, L24H28), again with linear $\alpha$-scaling
    and stable phase (\S\ref{sec:other}). Neither axis is numerical;
    both use the same extraction recipe as the gender compass, and the
    temporal plane is a genuine four-way dial with second-harmonic
    amplitude ratio $A_2/A_1 = 0.32$ (\S\ref{sec:harmonic}).
    \item \textbf{Downstream bias effects.} A single-site rank-$2$
    compass injection monotonically reduces WinoGender stereotypical
    co-reference correlation from $0.638\to 0.179$ on Phi-3 at $\alpha=1$
    with perplexity delta $+0.006$ (Table~\ref{tab:winogender}), and
    improves StereoSet gender ICAT by $4\text{--}6$ points at matched
    LMS (\S\ref{sec:stereoset}). The compass is therefore not only a
    diagnostic; it is an editing primitive with measurable,
    language-quality-neutral downstream effect.
    \item \textbf{Storage vs.\ steering regimes.} A rank-$2$ plane can
    be \emph{causally pushable} without being where the model \emph{stores}
    the category. GPT-2 hosts both regimes in adjacent layers (L9H7 steering,
    L10H9 storage); Gemma's L21H4 is storage; Phi-3's L24H10 is pure
    steering (\S\ref{sec:headshare}). The storage/steering distinction
    is a head-level property, not a model-level one.
    \item \textbf{Discussion of architecture differences.} We explain why
    polar-plot shapes differ across models (signal-to-noise on neutral
    prompts, singular-value scale, SVD sign ambiguity) without requiring
    different underlying mechanisms (\S\ref{sec:discussion}).
\end{enumerate}
