\section{Methods}
\label{sec:methods}

\subsection{Compass plane}
\label{sec:compass_plane}

For attention head $h$ at layer $\ell$, we compute the combined OV map
\begin{equation}
    W_{OV}^{(\ell,h)} = W_V^{(\ell,h)}\,W_O^{(\ell,h)}
    \in \mathbb{R}^{d_\text{model}\times d_\text{model}}
\end{equation}
and its singular value decomposition
$W_{OV}^{(\ell,h)} = U\,\Sigma\,V^\top$ with columns
$u_1,\dots,u_{d_\text{model}} \in \mathbb{R}^{d_\text{model}}$ and
singular values $\sigma_1 \geq \sigma_2 \geq \dots \geq 0$. A
\emph{compass plane} for this head is a pair of columns
$(u_i, u_j)$ with $i \neq j$; we refer to it as
$\mathcal{C}^{(\ell,h)}_{i,j} = \mathrm{span}(u_i, u_j)$.
The write-out directions $U$ are more natural than the read-in
directions $V$ for causal injection because they correspond to what
the head \emph{writes} into the residual stream.

A compass is \emph{semantic} when the vocabulary logits induced by unit
injections along the plane, $(W_U^\top u_i, W_U^\top u_j)$, exhibit a
consistent angular ordering of tokens that share a categorical axis (for
example, masculine / feminine pronouns at $0^\circ / 180^\circ$).

\subsection{Causal $\alpha$-sweep}
\label{sec:alpha_sweep}

The central test is a causal injection that varies both an \emph{angle}
$\theta$ and a \emph{scale} $\alpha$. Let
$v(\theta,\alpha) = \alpha\,\sigma_i\,\cos\theta\,u_i + \alpha\,\sigma_j\,\sin\theta\,u_j.$
For a given prompt $p$ whose final-token residual stream just before
layer $\ell$ is $r \in \mathbb{R}^{d_\text{model}}$, we run the forward
pass while replacing $r \gets r + v(\theta,\alpha)$ and record
$\mathrm{LD}(\theta,\alpha;p) = \mathrm{logit}_{t_{+}} - \mathrm{logit}_{t_{-}}$
for two antipodal probe tokens $(t_{+}, t_{-})$.
We average $\mathrm{LD}$ over three neutral prompts, sweep $\theta$ through
$36$ angles (step $10^\circ$) and $\alpha \in \{1,\,3,\,10\}$.

For each curve we fit a single sinusoid
$\mathrm{LD}(\theta) \approx \mu + A\cos(\theta - \phi)$
using the first DFT bin, and record mean $\mu$, amplitude $A$, phase $\phi$,
and the angles of maximum and minimum response.

\paragraph{Why amplitude--scaling + phase-stability is the key joint test.}
A random plane would produce neither effect. A plane that merely
correlates with the target without being read by the model would produce
some amplitude but no dependence on $\alpha$. A truly causal plane
produces amplitude that grows linearly with $\alpha$ (the intervention is
linear, so the logit-response is too, within the model's operating
range) and phase that is independent of $\alpha$ (the causal direction in
the plane does not depend on injection magnitude). We report both.

\subsection{Null and falsification tests}
\label{sec:nulls}

We complement the $\alpha$-sweep with a battery of tests implemented in
\texttt{investigate\_helix\_usage\_validated.py}:

\begin{itemize}
    \item \textbf{decode-heads.} Project $u_i, u_j$ through the unembedding
    $W_U$ and report the top tokens at $0^\circ, 90^\circ, 180^\circ, 270^\circ$.
    \item \textbf{cyclicity.} Verify that the polar angle of the compass
    projection wraps smoothly ($2\pi$-periodic) rather than saturating.
    \item \textbf{permutation.} Randomly permute the token set across the
    chosen antipodal pair and check that the compass signal vanishes.
    \item \textbf{random-plane.} Replace $(u_i, u_j)$ with a random pair
    of orthonormal directions drawn from $U(\mathrm{SO}(d_\text{model}))$
    and verify that the causal effect disappears.
    \item \textbf{principal-angles.} Compute principal angles between the
    compass plane and planes from unrelated heads; non-trivial overlap
    would indicate the compass is not head-specific.
    \item \textbf{scan-planes.} Enumerate $(i,j)$ pairs with $i,j \leq 10$
    and report the one with maximum causal amplitude; the target pair
    should be the global argmax.
    \item \textbf{semantic-ablate.} Zero out the projection
    $P_{\mathcal{C}}\,r$ and confirm that category-conditioned logits
    collapse toward neutral.
    \item \textbf{causal-patch.} Swap $P_{\mathcal{C}}\,r$ between a
    category-$+$ and a category-$-$ prompt and measure the resulting logit
    shift.
    \item \textbf{self-test.} End-to-end pipeline check on a fixed prompt
    to guard against implementation drift.
\end{itemize}

\subsection{Models and implementation}

We use TransformerLens \citep{nanda2022transformerlens} with
\texttt{HookedTransformer} on bfloat16. Hooks are installed at
\texttt{blocks.\{layer\}.hook\_resid\_pre}. Targets:

\begin{itemize}
    \item \textbf{GPT-2 Small} \citep{radford2019language}: gender L9H7
    $(u_1, u_2)$ (decode-discovered) and L10H9 $(u_0, u_3)$
    (scan-discovered).
    \item \textbf{Phi-3 Mini} \citep{abdin2024phi3}: gender L24H10
    $(u_0, u_1)$ (steering) and L28H1 $(u_0, u_3)$ (storage, late-layer);
    temporal L24H17 $(u_1, u_2)$; entity L24H28 $(u_3, u_7)$.
    \item \textbf{Gemma-2-2B} \citep{team2024gemma}: L21H4, $(u_0, u_2)$.
    \item \textbf{Llama-3.2-3B} \citep{grattafiori2024llama3}: L26H14,
    $(u_2, u_3)$ (scan-discovered).
\end{itemize}
