\section{Nine-test battery}
\label{app:battery}

Each compass passes the following (verdicts in parentheses are the
expected outcomes; see results for full logs):
\begin{itemize}
    \item \textbf{decode-heads} ($\checkmark$): $W_U^{\top}u_i$
    promotes readable categorical tokens at $0^\circ, 90^\circ,
    180^\circ, 270^\circ$.
    \item \textbf{cyclicity} ($\checkmark$): polar angle of the
    compass projection wraps smoothly, $2\pi$-periodic.
    \item \textbf{permutation null} ($\times$): shuffling the
    antipodal token set across prompts kills the compass signal.
    \item \textbf{random-plane null} ($\times$): random orthonormal
    directions from $U(\mathrm{SO}(d_\text{model}))$ remove the
    causal effect.
    \item \textbf{principal-angles} ($\checkmark$): fitted plane is
    not co-planar with unrelated heads.
    \item \textbf{scan-planes} ($\checkmark$): target $(i, j)$ is
    the argmax over $(i, j)$ pairs with $i, j \le 10$.
    \item \textbf{semantic-ablate} ($\times$): zeroing
    $P_{\mathcal{C}}\,r$ collapses category-conditioned logits to
    neutral.
    \item \textbf{causal-patch} ($\checkmark$): swapping
    $P_{\mathcal{C}}\,r$ between $\pm$-context prompts flips the
    logit sign.
    \item \textbf{self-test} ($\checkmark$): end-to-end pipeline
    reproducibility.
\end{itemize}
All four models pass all nine tests for the gender compass.

\section{StereoSet across domains}
\label{app:stereoset}

\begin{table}[h]
\centering
\small
\caption{StereoSet gender / race / profession at the compass
operating $\alpha$. Higher ICAT is better (LMS $\times$ fairness).
ActAdd is tuned to $\alpha=1$.}
\label{tab:stereoset}
\begin{tabular}{llcccc}
\toprule
Model (head) & Cond. / domain & LMS & SS & ICAT & $\Delta$ICAT \\
\midrule
\multirow{3}{*}{GPT-2 (L10H9)}
& baseline / gender      & 91.8 & 68.6 & 57.6 & --     \\
& compass 1.5 / gender   & 90.6 & 64.7 & \textbf{63.9} & $+6.4$ \\
& actadd / profession    & 88.7 & 59.0 & 72.7 & $+3.6$ \\
\midrule
\multirow{3}{*}{Phi-3 (L28H1)}
& baseline / gender      & 92.8 & 73.3 & 49.5 & --     \\
& compass 1.0 / gender   & 92.9 & 73.7 & 48.8 & $-0.7$ \\
& actadd / gender        & 92.8 & 69.8 & \textbf{56.0} & $+6.5$ \\
\midrule
\multirow{3}{*}{Gemma (L21H4)}
& baseline / gender      & 90.0 & 74.9 & 45.2 & --     \\
& compass 1.5 / gender   & 90.4 & 75.3 & 44.7 & $-0.5$ \\
& compass 1.0 / race     & 91.6 & 62.1 & 69.5 & $+0.2$ \\
\midrule
\multirow{3}{*}{Llama (L26H14)}
& baseline / gender      & 92.6 & 73.3 & 49.4 & --     \\
& compass 1.0 / gender   & 92.2 & 72.9 & \textbf{49.9} & $+0.5$ \\
& compass 1.5 / race     & 93.0 & 62.9 & 69.0 & $+0.7$ \\
\bottomrule
\end{tabular}
\end{table}

\section{Reproducibility}
\label{app:repro}

All experiments are reproducible from this repository. Seeds are
fixed in the pipeline; the $\alpha$-sweep is deterministic for
fixed prompts.

\begin{itemize}
    \item \texttt{experiments/compass\_causal\_sweep.py} --- causal
    $\alpha$-sweep and polar / curves / linearity figures.
    \item \texttt{investigate\_helix\_usage\_validated.py} ---
    nine-test battery.
    \item \texttt{src/models/online\_svd\_scanner.py} --- head survey
    and plane enumeration.
    \item \texttt{experiments/winogender\_eval.py} /
    \texttt{stereoset\_eval.py} --- downstream bias evaluation.
    \item \texttt{experiments/scan\_ci\_summary.py} --- Agresti-Coull
    CI aggregation for scan pass rates.
    \item \texttt{experiments/run\_llama\_suite.sh} --- end-to-end
    reproduction of Llama-3.2-3B scan and downstream runs.
\end{itemize}
